Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features

### 2022-03-04 publication/manuscript.tex
update arxiv preprint

\institute{
        Florian Pargent \orcidlink{0000-0002-2388-553X} \at
     Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstra{\ss}e 13, 80802 Munich, Germany \\
     \email{\href{mailto:florian.pargent@psy.lmu.de}{\nolinkurl{florian.pargent@psy.lmu.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Florian Pfisterer \orcidlink{0000-0001-8867-762X} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Janek Thomas \orcidlink{0000-0003-4511-6245} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Bernd Bischl \orcidlink{0000-0001-6002-6980} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstra{\ss}e 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

@@

While increasing sample size is usually considered the most important step to improve the predictive performance of a machine learning (ML) model, using effective feature engineering comes as a close second. One remaining challenge is how to handle high cardinality features -- categorical predictor variables with a high number of different levels but without any natural ordering.
While categorical variables with only a small number of possible levels can often be efficiently dealt with using standard techniques such as one-hot encoding, this approach becomes inefficient as the number of levels increases.
Despite this inefficiency, simpler strategies are often favored in practice because other methods are either not known, implementations are missing or because of a lack of trust due to missing validation studies.
Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML (Feurer et al., 2015; Thomas et al., 2018; Thornton et al., 2013). Optimally, strategies should be model-agnostic because benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
While a variety of strategies exist, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners (Bommert et al., 2020; Fernández-Delgado et al., 2014), as they can provide at least partial answers to such questions.
The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
Following calls in the computational statistics community for neutral benchmark studies (Boulesteix et al., 2017), which do not introduce a new method, thus reducing the risk of cherry picking methods (Dehghani et al., 2021) and reporting over-optimistic performance (Nie{\ss}l et al., 2021), we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.

@@

We broadly categorize feature encoding techniques into \emph{target-agnostic} methods and \emph{target-based} methods (Micci-Barreca, 2001).
Figure \ref{fig:overview} contains a taxonomy of our considered encoding methods. \emph{Target-agnostic} methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings.
Simple strategies from this domain e.g.~one-hot or dummy encoding are widely used -- in the scientific literature (Hancock and Khoshgoftaar, 2020; Kuhn and Johnson, 2019) but also on Kaggle\footnote{\url{https://www.kaggle.com}} to embed variables for classical ML algorithms as well as (deep) neural networks.
Such indicator methods map each level of a categorical variable to a set of dichotomous features encoding the presence or absence of a particular level.
An obvious drawback of indicator encoding is that it adds one additional feature per level of a categorical variable.
When indicator encoding leads to an unreasonable number of features, levels are often mapped to integer values with random order (integer encoding). Alternatively, the ``hashing trick'' (Weinberger et al., 2009) can be used to randomly collapse feature levels into a smaller number of indicator variables (Kuhn and Johnson, 2019), or levels can be encoded by using the observed frequency of a given level in the dataset (frequency encoding).

@@

\emph{Target-based} methods try to incorporate information about the target values associated with a given level.
Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature (for a brief mention, see Micci-Barreca, 2001).
The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value \(\hat{x}_l\) for the respective level.
An early formal description of this strategy is Micci-Barreca (2001).
In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used to encode that level for all observations: \(\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}\).
Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fails to generalize well for new observations.
In the extreme case of a categorical feature with unique values (e.g.~some hashed ID variable) studied in Prokhorenkova et al. (2018), the mean target for each level of this feature in simple target encoding is similar to the true target value of a single observation.
Based on the encoded feature, all observations can be predicted perfectly in the training set, even if the original variable did not contain any useful information.
ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
To avoid this, practitioners often use regularized target encoding with a smoothing parameter that shrinks those effects towards the global mean (Micci-Barreca, 2001).
An alternative strategy is to combine target encoding with cross-validation (CV) techniques (Prokhorenkova et al., 2018).

@@

Several small-scale studies have been previously conducted.
Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
One benchmark (\(6\) datasets) on encoding high cardinality features (cardinality between \(103\) and \(9095\)) in combination with gradient boosting has been published on the Kaggle forums (Prokopev, 2018).
Different versions of target encoding are compared with indicator, integer, and frequency encoding.
They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
Coors (2018) performed a benchmark (\(12\) datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library (Thomas et al., 2018). They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on \(2\) datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
A recent study (\(15\) datasets) found good performance of target encoding, but they only investigated categorical features in regression settings (Seca and Mendes-Moreira, 2021).
Several other publications studied encoding text data based on similarity (Cerda et al., 2018; Cerda and Varoquaux, 2020) and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
Wright and König (2019) study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between \(3\) and \(38\)), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
Prokhorenkova et al. (2018) compared their new CatBoost variant of target encoding to smoothed target encoding without CV, hold-out, and leave-one-out CV on \(8\) datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored towards neural networks (e.g.~the widely adopted \emph{entity embeddings} by Guo and Berkhahn, 2016) is provided in Hancock and Khoshgoftaar (2020).
They present a survey of indicator- and embedding-based methods but no benchmark study.
In contrast, our work is the first to focus on high cardinality variables and techniques that are agnostic to the subsequent ML method.
We study this problem on a larger variety of datasets and settings.
\newline

@@

Hash Encoding can be used to compute indicator variables based on a hash function (Weinberger et al., 2009).
The basic idea is to transform each feature level \(l\) into an integer \(hash(l) \in \mathbb{N}\), based on its label.
This integer is then transformed into an indicator representation, with \(1\) in indicator column number \((hash(l) \mod hash.size) + 1\) and \(0\) in all remaining columns (Kuhn and Johnson, 2019).
Some levels are hashed to the same indicator representation.
The smaller the \(hash.size\), the higher the number of collapsed levels.
The number of indicators is often effectively lower than \(hash.size\), as some indicators can be constant in the training set (we remove those columns for both training and prediction).
Although we could jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.

@@

Leaf encoding fits a decision tree on the training set to predict the target based on the categorical feature.
Each level is encoded by the number of the terminal node, in which an observation with the respective level ends up.
In that way, leaf encoding combines feature levels with similar target values.
We use the rpart package in R (Therneau and Atkinson, 2018) which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
Thus, our leaf encoder automatically uses an ``optimal'' number of new levels.
To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in Wright and König (2019).
New levels are encoded with the arbitrary number of the terminal node with most observations during training.
Encoded values are treated as a new categorical feature and encoded by one-hot encoding.
Our leaf encoder can be thought of as a simplification of the approach suggested by Grąbczewski and Jankowski (2003).

@@

An early formal description of a so-called impact, target or James-Stein encoder was provided by Micci-Barreca (2001).
The basic idea is to encode each feature level with the conditional target mean (regression) or the conditional relative frequency of one or more target classes (classification).
The impact encoder for classification uses a logit link and transforms the original feature into \(C\) numeric features, each representing one target class.
A smoothing parameter \(\epsilon\) is introduced to avoid division by zero.
This parameter could be used to further regularize towards the unconditional mean.
We choose a small \(\epsilon = 0.0001\) as we want to compare simple target encoding with the regularized encoder introduced next.
Weights of evidence encoding from the credit scoring classification literature (Hand and Henley, 1997) is almost identical to impact encoding, but without regularization or centering.

@@

Smoothed target encoding (Micci-Barreca, 2001) can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in Kuhn and Johnson (2019).
To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification.
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set Bates (2020).
The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
In multiclass classification, we fit \(C\) one vs.~rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

Overfitting can be further avoided through combination with cross-validation (CV) to train the encoder on independent observations without limiting the data to train the ML model.
We provide an implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into \(n.folds\) and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the \(n.folds\) CV models (for \(n.folds\) \textgreater{} 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with \(5-\) (5CV) and with \(10-\) fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R (Bates et al., 2015) as an efficient way to fit glmms.

@@

We include three control conditions to better understand the effectiveness of the investigated encoders:
The performance of a featureless learner (\textbf{FL} condition) was estimated as a conservative baseline for each dataset.
In regression problems, FL predicts the mean of the target variable in the training set for each observation in the test set.
In classification problems, the most frequent class of the target within the training set is predicted.
For each dataset, we also consider a RF without encoding (\textbf{none} condition), to compare the use of encoding methods with a natural categorical splitting approach.
The ranger (Wright and Ziegler, 2017) implementation provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm (Wright and König, 2019).
In the \textbf{remove} high cardinality features control condition, we omit features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
This condition reflects on whether including high cardinality features does indeed improve predictive performance.
Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
We include an overview of available implementations in widely used ML frameworks for R and python in the Supplementary Material.

@@

A table showing a detailed summary of all benchmark datasets can be found in the Supplementary Material.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies (Cerda et al., 2018; Coors, 2018; Kuhn and Johnson, 2019; Prokhorenkova et al., 2018).
All datasets can be downloaded from the \texttt{OpenML} platform (Vanschoren et al., 2013) based on the name or the displayed OmlId.
The datasets include 8 regression, 10 binary classification, and 6 multiclass classification problems (between 3 and 12 classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable.
The maximum normalized entropy is \(1\), which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between 736 and 1224158 observations.
The total number of features ranges between 5 and 208.
Datasets contain between 1 and 20 categorical features with more than \(10\) levels, with the maximum number of levels for a feature ranging between 14 and 30114. Missing values are present in about half of the datasets.

@@

It is often assumed that advanced encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
To reflect this in our benchmark, a high cardinality threshold (HCT) parameter with values of \(10\), \(25\), and \(125\) was introduced determining varying configurations for the different encoders.
For indicator encoding, the \(HCT - 1\) most frequent levels are encoded together with a single collapsed category for the remaining levels.
For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded. HCT is used as the hash size in hash encoding.
In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
This ensured that encoders always affect at least one feature and the remove condition always removes at least one feature.
If settings where aan encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}

@@

In total, we investigate \(5\) ML algorithms. We kept tuning their hyperparameters to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of ML algorithms.
LASSOs were fitted with \texttt{glmnet} (Friedman et al., 2010), internally tuning the regularization using 5-fold CV.
RFs with \(500\) trees were trained using \texttt{ranger} (Wright and Ziegler, 2017) without tuning, because RFs can be expected to give reasonable results with default settings (Probst et al., 2019).
GB models were trained using \texttt{xgboost} (Chen et al., 2018), setting the learning rate to \(0.01\) and determining the number of iterations using early stopping on a 20\% holdout set.
KNN was taken from package \texttt{kknn} (Schliep and Hechenbichler, 2016) standardizing features and using a constant \(k=15\) for the number of nearest neighbors, together with an information gain filter (Brown et al., 2012) to limit the number of features to \(25\).
SVMs with radial basis function kernel were trained with \texttt{liquidSVM} (Steinwart and Thomann, 2017).
The bandwidth and regularization parameters were internally tuned using 5-fold CV. We used a one-vs-all approach for multiclass-settings.

@@

We perform all analyses in the open-source statistical software R (R Core Team, 2021).
To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the \texttt{mlrCPO} package (Binder, 2018).
The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the \texttt{mlr} framework (Bischl et al., 2016) together with the \texttt{batchtools} package (Lang et al., 2017) to scale the benchmark analysis to HPC compute infrastructure.
All materials for this study, including reproducible code for this manuscript and result objects can be downloaded from our \textbf{online repository} \footnote{\url{https://github.com/slds-lmu/paper_2021_categorical_feature_encodings}}.

Throughout our experiments, we use 5-fold CV to obtain estimates of predictive performance.
Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU (Ferri et al., 2009) for multiclass problems.

@@

To further analyze our results, we used statistical inference methods inspired by the benchmark community in computational statistics (e.g., Hothorn et al., 2005).
First, we present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}:
We defined an encoder relation within each dataset, based on corrected resample t-tests (Nadeau and Bengio, 2003).
An encoding was defined to beat another encoding if the one-sided p-value of the t-test was \(< .05\).
This allowed us to compute a weak-order consensus ranking \(R\) defined by the optimization problem:

@@

where \(d\) is the symmetric difference distance and \(R_b\) is the relation for dataset \(b\) (Hornik and Meyer, 2007; Meyer and Hornik, 2018).
The symmetric difference between two relations is the number of cases one encoding beats another encoding in one relation but not in the other one.

@@

\bibliographystyle{spbasic}

>>>>

\institute{
        Florian Pargent \orcidlink{0000-0002-2388-553X} \at
     Department of Psychology, Psychological Methods and Assessment, LMU Munich, Leopoldstraße 13, 80802 Munich, Germany \\
     \email{\href{mailto:florian.pargent@psy.lmu.de}{\nolinkurl{florian.pargent@psy.lmu.de}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Florian Pfisterer \orcidlink{0000-0001-8867-762X} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Janek Thomas \orcidlink{0000-0003-4511-6245} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Bernd Bischl \orcidlink{0000-0001-6002-6980} \at
     Department of Statistics, Statistical Learning and Data Science, LMU Munich, Ludwigstraße 33, 80539 Munich, Germany \\
     %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

\date{\textbf{This preprint has since been published in Computational Statistics:} \newline Pargent, F., Pfisterer, F., Thomas, J., \& Bischl, B. Regularized target encoding outperforms traditional methods in supervised machine learning with high cardinality features. \textit{Computational Statistics} (2022). https://doi.org/10.1007/s00180-022-01207-6 \newline \textbf{Link to copy of record:} https://link.springer.com/article/10.1007/s00180-022-01207-6}

@@

While increasing sample size is usually considered the most important step to improve the predictive performance of a machine learning (ML) model, using effective feature engineering comes as a close second. One remaining challenge is how to handle high cardinality features -- categorical predictor variables with a high number of different levels but without any natural ordering.
While categorical variables with only a small number of possible levels can often be efficiently dealt with using standard techniques such as one-hot encoding, this approach becomes inefficient as the number of levels increases.
Despite this inefficiency, simpler strategies are often favored in practice because other methods are either not known, implementations are missing or because of a lack of trust due to missing validation studies.
Although domain knowledge can sometimes be used to reduce the number of theoretically relevant levels, finding strategies that work well on a large variety of problems is highly important for many applications as well as in automated ML (Thomas, Coors, and Bischl 2018; Thornton et al. 2013; Feurer et al. 2015). Optimally, strategies should be model-agnostic because benchmarking encoding methods together with ML algorithms from different classes is often necessary for applications.
While a variety of strategies exist, there are very few benchmarks that can be used to decide which technique is expected to yield good predictive performance.
Furthermore, there has recently been increasing attention on scientific benchmark studies that compare different methods to provide a clearer picture in light of a large number of methods available to practitioners (Fernández-Delgado et al. 2014; Bommert et al. 2020), as they can provide at least partial answers to such questions.
The goal of this study is to provide an overview of existing approaches for encoding categorical predictor variables and to study their effect on a model's predictive performance.
Following calls in the computational statistics community for neutral benchmark studies (Boulesteix et al. 2017), which do not introduce a new method, thus reducing the risk of cherry picking methods (Dehghani et al. 2021) and reporting over-optimistic performance (Nießl et al., n.d.), we present a carefully designed experimental setting to discern the effect of encoding strategies and their interaction with different ML algorithms.

@@

We broadly categorize feature encoding techniques into \emph{target-agnostic} methods and \emph{target-based} methods (Micci-Barreca 2001).
Figure \ref{fig:overview} contains a taxonomy of our considered encoding methods. \emph{Target-agnostic} methods do not rely on any information about the target variable and can therefore also be used in unsupervised settings.
Simple strategies from this domain e.g.~one-hot or dummy encoding are widely used -- in the scientific literature (Kuhn and Johnson 2019; Hancock and Khoshgoftaar 2020) but also on Kaggle\footnote{\url{https://www.kaggle.com}} to embed variables for classical ML algorithms as well as (deep) neural networks.
Such indicator methods map each level of a categorical variable to a set of dichotomous features encoding the presence or absence of a particular level.
An obvious drawback of indicator encoding is that it adds one additional feature per level of a categorical variable.
When indicator encoding leads to an unreasonable number of features, levels are often mapped to integer values with random order (integer encoding). Alternatively, the ``hashing trick'' (Weinberger et al. 2009) can be used to randomly collapse feature levels into a smaller number of indicator variables (Kuhn and Johnson 2019), or levels can be encoded by using the observed frequency of a given level in the dataset (frequency encoding).

@@

\emph{Target-based} methods try to incorporate information about the target values associated with a given level.
Early strategies aimed to reduce the number of levels by methods like hierarchical clustering or decision trees based on statistics of the target variable, although this has been rarely described in the scientific literature (for a brief mention, see Micci-Barreca 2001).
The basic idea of more advanced methods called target, impact, mean, or likelihood encoding is to use the training set to make a simple prediction of the target for each level of the categorical feature, and to use the prediction as the numerical feature value \(\hat{x}_l\) for the respective level.
An early formal description of this strategy is Micci-Barreca (2001).
In simple target encoding for regression problems, the mean target value in the training set from all observations with a certain feature level is used to encode that level for all observations: \(\hat{x}_l = \frac{\sum_{i:x^{train}_i = l}y^{train}_i}{N_l}\).
Simple target encoding often does not perform well with rare levels, where it tends to overfit to the training data and fails to generalize well for new observations.
In the extreme case of a categorical feature with unique values (e.g.~some hashed ID variable) studied in Prokhorenkova et al. (2018), the mean target for each level of this feature in simple target encoding is similar to the true target value of a single observation.
Based on the encoded feature, all observations can be predicted perfectly in the training set, even if the original variable did not contain any useful information.
ML models would place a high priority on such an encoded feature during training but would perform badly on test data.
To avoid this, practitioners often use regularized target encoding with a smoothing parameter that shrinks those effects towards the global mean (Micci-Barreca 2001).
An alternative strategy is to combine target encoding with cross-validation (CV) techniques (Prokhorenkova et al. 2018).

@@

Several small-scale studies have been previously conducted.
Those works did not yield conclusive results due to narrower scopes or not considering high cardinality variables.
One benchmark (\(6\) datasets) on encoding high cardinality features (cardinality between \(103\) and \(9095\)) in combination with gradient boosting has been published on the Kaggle forums (Prokopev 2018).
Different versions of target encoding are compared with indicator, integer, and frequency encoding.
They recommend combining the smoothed version of target encoding with 4- or 5-fold CV, never using simple target encoding and using indicator encoding only for small datasets.
Interestingly, frequency encoding did perform well in many cases.
Coors (2018) performed a benchmark (\(12\) datasets) on encoding high cardinality features (maximum number of levels per dataset between 10 and 25847) when developing the automatic gradient boosting (autoxgboost) library (Thomas, Coors, and Bischl 2018). They compared different variants of target encoding with integer and indicator encoding. In their benchmarks, target encoding only improved over target-agnostic methods on \(2\) datasets, while it led to worse results on 4 datasets. For a smaller number of levels, indicator and integer encoding yielded similar results.
As those studies only consider gradient boosting and a limited amount of datasets, it is unclear whether results generalize to other datasets and ML algorithms.
A recent study (\(15\) datasets) found good performance of target encoding, but they only investigated categorical features in regression settings (Seca and Mendes-Moreira 2021).
Several other publications studied encoding text data based on similarity (Patricio Cerda, Varoquaux, and Kégl 2018; P. Cerda and Varoquaux 2020) and employed indicator or target encoding as baselines.
Another line of work studies variable encodings employed within specific ML models.
Wright and König (2019) study treatments for categorical variables in random forests together with dummy and integer encoding (18 datasets, cardinality between \(3\) and \(38\)), concluding that indicator and integer encoding perform subpar in comparison to methods that re-order levels according to the target variable.
Prokhorenkova et al. (2018) compared their new CatBoost variant of target encoding to smoothed target encoding without CV, hold-out, and leave-one-out CV on \(8\) datasets.
While the CatBoost method performed best, hold-out came second and target encoding without CV performed worst.
An overview of encoding techniques tailored towards neural networks (e.g.~the widely adopted \emph{entity embeddings} by Guo and Berkhahn 2016) is provided in Hancock and Khoshgoftaar (2020).
They present a survey of indicator- and embedding-based methods but no benchmark study.
In contrast, our work is the first to focus on high cardinality variables and techniques that are agnostic to the subsequent ML method.
We study this problem on a larger variety of datasets and settings.
\newline

@@

Hash Encoding can be used to compute indicator variables based on a hash function (Weinberger et al. 2009).
The basic idea is to transform each feature level \(l\) into an integer \(hash(l) \in \mathbb{N}\), based on its label.
This integer is then transformed into an indicator representation, with \(1\) in indicator column number \((hash(l) \mod hash.size) + 1\) and \(0\) in all remaining columns (Kuhn and Johnson 2019).
Some levels are hashed to the same indicator representation.
The smaller the \(hash.size\), the higher the number of collapsed levels.
The number of indicators is often effectively lower than \(hash.size\), as some indicators can be constant in the training set (we remove those columns for both training and prediction).
Although we could jointly hash multiple features, we hash each feature separately to improve comparability with the other encoders.

@@

Leaf encoding fits a decision tree on the training set to predict the target based on the categorical feature.
Each level is encoded by the number of the terminal node, in which an observation with the respective level ends up.
In that way, leaf encoding combines feature levels with similar target values.
We use the rpart package in R (Therneau and Atkinson 2018) which grows CARTs with categorical feature support that can be pruned based on internal performance estimates from 10-fold CV.
Thus, our leaf encoder automatically uses an ``optimal'' number of new levels.
To speed up the computation for multiclass classification, our implementation uses the ordering approach presented in Wright and König (2019).
New levels are encoded with the arbitrary number of the terminal node with most observations during training.
Encoded values are treated as a new categorical feature and encoded by one-hot encoding.
Our leaf encoder can be thought of as a simplification of the approach suggested by Grąbczewski and Jankowski (2003).

@@

An early formal description of a so-called impact, target or James-Stein encoder was provided by Micci-Barreca (2001).
The basic idea is to encode each feature level with the conditional target mean (regression) or the conditional relative frequency of one or more target classes (classification).
The impact encoder for classification uses a logit link and transforms the original feature into \(C\) numeric features, each representing one target class.
A smoothing parameter \(\epsilon\) is introduced to avoid division by zero.
This parameter could be used to further regularize towards the unconditional mean.
We choose a small \(\epsilon = 0.0001\) as we want to compare simple target encoding with the regularized encoder introduced next.
Weights of evidence encoding from the credit scoring classification literature (Hand and Henley 1997) is almost identical to impact encoding, but without regularization or centering.

@@

Smoothed target encoding (Micci-Barreca 2001) can be interpreted as a simple (generalized) linear mixed model (glmm) in which the target is predicted by a random intercept for each feature level in addition to a fixed global intercept.
This connection is described in Kuhn and Johnson (2019).
To achieve regularized impact encoding, we implemented glmm encoders for regression, binary, and multiclass classification.
The encoded value for each level is based on the spherical conditional mode estimates.
In regression, the conditional modes are similar to the mean target value for each level, weighted by the relative observed frequency of that level in the training set (Gelman and Hill 2006; Bates 2020).
The estimate of the fixed intercept can be used during the prediction phase to encode new feature levels not observed in the training set.
In multiclass classification, we fit \(C\) one vs.~rest glmms resulting in one encoded feature per class.
An important advantage of using a glmm over impact encoding with a smoothing parameter is that a reasonable amount of regularization is determined automatically and tuning the complete ML pipeline is not necessary.

Overfitting can be further avoided through combination with cross-validation (CV) to train the encoder on independent observations without limiting the data to train the ML model.
We provide an implementation which combines target encoding based on glmms with CV.
During the training phase, we partition the data using CV into \(n.folds\) and fit a glmm on each resulting training set.
For each observation, there is exactly one glmm that did not use that observation for model fitting and can be safely used for encoding.
Note that the \(n.folds\) CV models (for \(n.folds\) \textgreater{} 1) are only used during the training phase.
In the prediction phase, feature values are always encoded by a single glmm fitted to the complete training set.
We study this method in three different settings: without CV (noCV), with \(5-\) (5CV) and with \(10-\) fold CV (10CV).
In our study, we use the lmer (regression) and glmer (classification) functions from the lme4 package in R (Bates et al. 2015) as an efficient way to fit glmms.

@@

We include three control conditions to better understand the effectiveness of the investigated encoders:
The performance of a featureless learner (\textbf{FL} condition) was estimated as a conservative baseline for each dataset.
In regression problems, FL predicts the mean of the target variable in the training set for each observation in the test set.
In classification problems, the most frequent class of the target within the training set is predicted.
For each dataset, we also consider a RF without encoding (\textbf{none} condition), to compare the use of encoding methods with a natural categorical splitting approach.
The ranger (Wright and Ziegler 2017) implementation provides efficient categorical feature support by ordering levels once before starting the tree growing algorithm (Wright and König 2019).
In the \textbf{remove} high cardinality features control condition, we omit features with a high number of levels above some threshold and use one-hot encoding (without collapsing rare levels) for the remaining features.
This condition reflects on whether including high cardinality features does indeed improve predictive performance.
Otherwise, the best encoding might just provide the least impairment compared to not including any high cardinality features.
We include an overview of available implementations in widely used ML frameworks for R and python in the Supplementary Material.

@@

A table showing a detailed summary of all benchmark datasets can be found in the Supplementary Material.
We specifically investigate datasets that contain categorical variables with a large number of levels, including many well-known datasets
from previous studies (Patricio Cerda, Varoquaux, and Kégl 2018; Prokhorenkova et al. 2018; Coors 2018; Kuhn and Johnson 2019).
All datasets can be downloaded from the \texttt{OpenML} platform (Vanschoren et al. 2013) based on the name or the displayed OmlId.
The datasets include 8 regression, 10 binary classification, and 6 multiclass classification problems (between 3 and 12 classes).
To assess the imbalance in categorical variables we computed the normalized entropy for each categorical variable.
The maximum normalized entropy is \(1\), which corresponds to a uniform distribution, while a lower number indicates a larger imbalance.
Sample sizes range between 736 and 1224158 observations.
The total number of features ranges between 5 and 208.
Datasets contain between 1 and 20 categorical features with more than \(10\) levels, with the maximum number of levels for a feature ranging between 14 and 30114. Missing values are present in about half of the datasets.

@@

It is often assumed that advanced encoding methods are only advantageous for variables with a high number of levels, while simple indicator encoding is more appropriate for a small number of levels.
To reflect this in our benchmark, a high cardinality threshold (HCT) parameter with values of \(10\), \(25\), and \(125\) was introduced determining varying configurations for the different encoders.
For indicator encoding, the \(HCT - 1\) most frequent levels are encoded together with a single collapsed category for the remaining levels.
For integer, frequency, hash, leaf, impact, and glmm encoders, only features with more than HCT levels in the training set were encoded with the respective strategy, while the remaining categorical features were one-hot encoded. HCT is used as the hash size in hash encoding.
In the remove control condition, features with more than HCT levels in the training set are removed from the feature set.\footnote{Based on the number of categorical features and levels per feature, some HCT settings were removed from the benchmark for some combinations of dataset $\times$ encoder.
This ensured that encoders always affect at least one feature and that the remove condition always removes at least one feature.
If settings where an encoder would lead to identical encoding strategies for all features of a dataset, we only kept the condition with the smallest HCT value.}

@@

In total, we investigate \(5\) ML algorithms. We kept tuning their hyperparameters to a minimum because we were interested in the effect of the encoding techniques instead of a comparison of ML algorithms.
LASSOs were fitted with \texttt{glmnet} (Friedman, Hastie, and Tibshirani 2010), internally tuning the regularization using 5-fold CV.
RFs with \(500\) trees were trained using \texttt{ranger} (Wright and Ziegler 2017) without tuning, because RFs can be expected to give reasonable results with default settings (Probst, Wright, and Boulesteix 2019).
GB models were trained using \texttt{xgboost} (Chen et al. 2018), setting the learning rate to \(0.01\) and determining the number of iterations using early stopping on a 20\% holdout set.
KNN was taken from package \texttt{kknn} (Schliep and Hechenbichler 2016) standardizing features and using a constant \(k=15\) for the number of nearest neighbors, together with an information gain filter (Brown et al. 2012) to limit the number of features to \(25\).
SVMs with radial basis function kernel were trained with \texttt{liquidSVM} (Steinwart and Thomann 2017).
The bandwidth and regularization parameters were internally tuned using 5-fold CV. We used a one-vs-all approach for multiclass-settings.

@@

We perform all analyses in the open-source statistical software R (R Core Team 2021).
To enable a fair and reliable comparison in our study, we implemented all encoding methods on top of the \texttt{mlrCPO} package (Binder 2018).
The pre-processing, as well as the final ML algorithm described in \ref{sec:pipeline}, were trained and resampled using the \texttt{mlr} framework (Bischl et al. 2016) together with the \texttt{batchtools} package (Lang, Bischl, and Surmann 2017) to scale the benchmark analysis to HPC compute infrastructure.
All materials for this study, including reproducible code for this manuscript and result objects can be downloaded from our \textbf{online repository} \footnote{\url{https://github.com/slds-lmu/paper_2021_categorical_feature_encodings}}.

Throughout our experiments, we use 5-fold CV to obtain estimates of predictive performance.
Depending on the target variable, we report root mean squared error (RMSE) for regression, area under the curve (AUC) for binary classification, and its extension AUNU (Ferri, Hernández-Orallo, and Modroiu 2009) for multiclass problems.

@@

To further analyze our results, we used statistical inference methods inspired by the benchmark community in computational statistics (e.g., Hothorn et al. 2005).
First, we present meta rankings for each ML algorithm in Figure \ref{fig:rank-res}:
We defined an encoder relation within each dataset, based on corrected resample t-tests (Nadeau and Bengio 2003).
An encoding was defined to beat another encoding if the one-sided p-value of the t-test was \(< .05\).
This allowed us to compute a weak-order consensus ranking \(R\) defined by the optimization problem:

@@

where \(d\) is the symmetric difference distance and \(R_b\) is the relation for dataset \(b\) (Hornik and Meyer 2007; Meyer and Hornik 2018).
The symmetric difference between two relations is the number of cases one encoding beats another encoding in one relation but not in the other one.

@@

\bibliographystyle{spbasic}
\bibliography{literature.bib}
